A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-13 article/03_Results.Rmd
language in the results chapter

This narrative is wrong or inaccurate in some aspects. Please see the analysis in Section \ref{time-series-discussion}.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared euclidian distance for each region-region relationship. The majority of the curves in its upper right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the fact that the transition process in each region as depicted in Radon-B is different and that it took place with a time lag compared to other regions. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux region, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is also due to how correctly or incorrectly the proxy developed from radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

Table \ref{tab:distanceandcorrelationtable*} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time window at the transition to the Middle Bronze Age from 1800 to 1600calBC is the 5% significance level of the mantle test almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED network based on *burial type* and the spatial distance several observations result from Table \ref{tab:distanceandcorrelationtable*}: Surprisingly, cultural distance does not seem to correlate with spatial distance. A significant correlation between the distance measures only reveals itself for the Early Bronze Age time window 2000 to 1800calBC: Only in the Early Bronze Age, long before the universal switch to the Urnfield burial, large regions directly adjacent to each other tend to undergo the same developments rather than the ones further away from each other. In the Middle Bronze Age, the SED developments of the *burial type* variable undergo the peaks already observed in Figure \ref{sed_region_matrix}. The correlation between cultural and spatial distance over these peaks is low and even tilts between 1400 and 1200calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction* (Table \ref{tab:distanceandcorrelationtable*}), similar conclusions can be drawn as in the *burial type* context: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated, but here, too, the early bronze age is by far the phase of the most substantial spatial correlation. It is only significant for the time window between 1800 and 1400calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.

>>>>

This narrative is wrong or inaccurate in some aspects. Please see the analysis in section \ref{time-series-discussion}.

@@

Beyond the simple comparison of archaeological knowledge and information derived from Radon-B, Figure \ref{sed_region_matrix} shows the time series of squared euclidian distance for each region-region relationship. The majority of the curves in its top right half for *burial type* correspond to expectations: At the beginning of the Bronze Age, inhumation is widespread throughout the study area, so the Euclidean distance at 2200calBC is small in almost every region-region relationship. Only the Benelux region, where the share of cremation is already high in the Early Bronze Age, is excluded. From the Middle Bronze Age onwards, but especially at the transition to the Late Bronze Age, the transition to the dominance of cremation took place all over Europe. The SED values increase in this period and show clear peaks or several successive maxima. This rise is a consequence of the time lag in the transition processes as depicted in Radon-B: It is different in every region. At the end of the Late Bronze Age, the average SED is approaching zero again, as burial by cremation dominates in almost all of Europe. The early and complete adoption of the cremation rite in the Benelux area, as depicted in Radon-B, clearly sets it apart from the other regions. England, on the other hand, with its slow, gradual transition from inhumation to cremation, stands out due to its consistently moderate distances to all other regions. It remains without accentuated minima and maxima. The situation for *burial construction* is more complicated, as can be seen from the plots in the bottom right corner of Figure \ref{sed_region_matrix}. The distance does not seem to follow a general trend, but a multitude of local and temporally limited phenomena. This heterogeneity is of course also due to how correctly or incorrectly the proxy developed from Radon-B depicts developments in the real world. In the Middle Bronze Age, the Tumulus culture should bring about a fundamental reduction of the SED for almost all region-region relationships -- this effect is indeed weakly visible.

Table \ref{tab:distanceandcorrelationtable*} shows that the distance networks of *burial type* and *burial construction* are not equivalent. There is also no significant negative correlation. The degree of agreement between two regions in the question of inhumation and cremation is therefore not suitable for predicting the similarity of developments in these regions concerning the dominance of flat or mound graves. Only for the time window at the transition to the Middle Bronze Age from 1800 to 1600calBC is the 5% significance level of the Mantle test almost undercut, thus suggesting a statistically significant relationship.

For the relationship between the SED network based on *burial type* and the spatial distance several observations result from Table \ref{tab:distanceandcorrelationtable*}: Cultural distance calculated from these to variables does not seem to correlate with spatial distance. A significant correlation between the distance measures only reveals itself for the Early Bronze Age time slot 2000 to 1800calBC: Only in the Early Bronze Age, long before the universal switch to the Urnfield burial tradition, large regions directly adjacent to each other tend to undergo the same developments rather than the ones further away from each other. In the Middle Bronze Age the SED developments of the *burial type* variable undergo the peaks already observed in Figure \ref{sed_region_matrix}. The correlation between cultural and spatial distance over these peaks is low and even tilts between 1400 and 1200calBC, at the transition to the Late Bronze Age, into the negative -- although not significantly. The cultural distance between several geographically distant regions is then smaller than between several directly adjacent regions. Looking at the SED network of *burial construction* (Table \ref{tab:distanceandcorrelationtable*}), similar conclusions can be drawn as in the *burial type* context: The distance values here are larger overall, and the SED differences within the individual spatial distance classes are clearly more accentuated, but here, too, the early bronze age is by far the phase of the most substantial spatial correlation. It is only significant for the time slot between 1800 and 1400calBC. There is no time period with negative mean correlation for *burial construction* cultural distance and spatial distance, but also here the correlation coefficient approaches zero in the Late Bronze Age.
